First of all, I would like to thank my PhD advisor, Noah Goodman. When I was choosing which graduate school to attend, one of my mentors said something along the lines of ``Go find Noah Goodman. He is a genius.''. Over five years, I have gotten to learn and re-learn what being a genius means. Noah's mentorship and guidance has been the principle component of what made my learning and my contributions possible. Thank you for teaching me how to be curious and how to search for the answers no matter how complex the problem. These are lessons that that I will take with me for many years to come.

The previous and current members of the ``CoCoLab'' have been indispensible collaborators and friends. Judy Fan and Robert Hawkins were early role models for me: two shining examples of what being stellar researchers is like; Ben Peloquin and Sahil Chopra were partners in crime. Thank you for helping me grow as a researcher and for all the fun we had along the way (I think back to those days we played basketball together with fond memories). As I slowly became a more senior student in CoCoLab, I had the priviledge of meeting more talented researchers. Alex Tamkin, Jesse Mu, and Rose Wang inspire me with their research vision and process. A special thank you to Rose for instilling new research drive in me in my final year. Outside of CoCoLab, Kristy Choi, Shyamal Buch, and Ali Malik have been pillars of my personal community at Stanford. Thank you for spending time with me. Whether it was tennis, meals, or talking shop, your friendships have enriched me academically and beyond.

I am greatly in debt to a set of outstanding mentors that have helped me to and through my PhD. So many times was I underqualified and out of my depth. So many times have others have taken leaps of faith and given me an opportunity. To Brendan Morris for teaching me about optical flow, my first ``AI'' algorithm. To Finale Doshi-Velez, Marzyeh Ghassemi, and Michael Hughes, with whom I worked on my first research project with. To Yura Perov and Frank Wood, who first taught me what inference was. To Manohar Paluri, Nikhil Johri, and the Facebook Lumos team for showing me the practical applications of AI. One of my favorite parts of the PhD was collaborating with other faculty. Thank you to Chris Piech, Stefano Ermon, Chelsea Finn, Ben Domingue, Dan Yamins, Andrew Maas, John Mitchell for the many hours toiling over experiments, checking proofs, drawing and re-drawing (and then re-drawing again) figures. Your collective guidance and advice has shaped my research mindset. I have also enjoyed mentoring students who have taught me so much in return: Oliver Zhang, Milan Mosse, Hokyung Sung, Conner Vercellino, Ron Arel, Jasmine Bayrooti, Ananya Karthik, John Nguyen, Vrinda Vasavada, and Will Song.

On a more personal level, thank you to the past and current members of Bawksquad, my roommates during my graduate studies, for being my second family. I am forever indebted to ny parents, Nick and Wei, for their unconditional love and supported. They are the reason I wanted to pursue a PhD and the reason for my love of learning.
Finally, I would like to thank my significant other, Connie Liu. For all that you do, I would like to think this dissertion is partly written by you. Over the last decade, you have continued to push me, to support me, and to inspire me. To the next chapter.

% Mike Wu is supported by NSF-GRFP and SIGF as the Karr Family Fellow.